---
title: "Brazilian Portuguese Nonce-Word Acceptability Analysis"
author:
  - David Nguyen
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 8,
  fig.height = 5
)
# Install & load required packages
pkgs <- c("ggplot2","ggmosaic")
for(pkg in pkgs) {
  if (!requireNamespace(pkg, quietly=TRUE)) install.packages(pkg)
}
lapply(pkgs, library, character.only=TRUE)
````

# 1. Project Overview

We investigate whether **word structure**—specifically **syllable count** (length) and **initial segment type**—influences acceptability judgments in a nonce-word experiment with native Brazilian Portuguese speakers.

* **Null hypothesis (H₀):** Word structure does *not* influence acceptability judgments.
* **Alternative hypothesis (H₁):** Word structure *does* influence acceptability judgments.

# 2. Data Loading

```{r load-data}
bp <- read.csv("bp-nonce.csv", stringsAsFactors = TRUE)
cat("Total observations:", nrow(bp), "\n\n")
head(bp)
```

The dataset `bp-nonce.csv` contains:

* `response`: whether the nonce word was **accepted** or **rejected**
* `length`: the number of syllables (e.g., `"1"`, `"2"`)
* `initial`: type of initial segment (e.g., `"stop"`, `"fricative"`, etc.)

# 3. Contingency Tables

Summarize counts of responses by `length` and `initial`:

```{r summary-table}
summary_table <- xtabs(~ response + length + initial, data = bp)
print(summary_table)
```

# 4. Mosaic Plot Visualization

We visualize the joint distribution of `response` × `length` × `initial` using a faceted mosaic plot:

```{r mosaic-plot}
ggplot(bp) +
  geom_mosaic(aes(weight = 1, x = product(response), fill = response)) +
  facet_grid(length ~ initial) +
  labs(
    title = "Mosaic Plot of Acceptability by Syllable Count and Initial Segment",
    x     = "Response",
    y     = "Proportion"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    panel.grid  = element_blank()
  )
```

# 5. Chi-Square Tests

## 5.1 Response vs. Initial Segment

Test whether **initial segment type** is associated with **acceptability**:

```{r chi-response-initial}
cs_initial <- chisq.test(xtabs(~ response + initial, data = bp))
cs_initial
```

> **Result:** p-value ≈ `r signif(cs_initial$p.value,3)`
> At α = 0.05, we **fail to reject** H₀: *no significant association* between initial segment and response.

## 5.2 Length vs. Initial Segment

Test independence of **syllable count** and **initial segment**:

```{r chi-length-initial}
cs_length <- chisq.test(xtabs(~ length + initial, data = bp))
cs_length
```

> **Result:** p-value = `r signif(cs_length$p.value,3)`
> As expected, we observe **no association** (variables are independent by design).

# 6. Interpretation

* The chi-square test for **response vs. initial** yields p > 0.05, indicating **no evidence** that initial segment influences acceptability judgments in this sample.
* The mosaic plot confirms a fairly uniform pattern across facets.
* The **syllable count** and **initial segment** are also independent, validating experimental design.

**Conclusion:** We find no support for the alternative hypothesis; word structure (initial segment type) does *not* significantly influence nonce-word acceptability among our participants.

# 7. Next Steps

* Test **response vs. length** directly, or include both predictors in a **logistic regression**:

  ```r
  glm(response ~ length + initial, family="binomial", data=bp)
  ```
* Explore additional phonological variables (e.g., coda type, stress pattern).
* Increase sample size or participant diversity to improve statistical power.

# 8. References

* **ggplot2**: Wickham H. (2016). *ggplot2: Elegant Graphics for Data Analysis*.
* **ggmosaic**: Hofmann H. (2016). *Visualizing Categorical Data with Mosaic Plots*.
* Brazilian Portuguese nonce-word experiment (unpublished data).
